---
title: "Countries"
author: "Samir Gadkari"
date: "3/1/2021"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rjson)
library(RPostgres)
library(DBI)
library(readxl)
```


We will explore which countries are getting better with a view towards
investing in them.

Data obtained from: 

  * https://www.resourcepanel.org/global-material-flows-database
  * https://github.com/owid/owid-datasets which gets it's data from
    https://ourworldindata.org/global-education

Ideas to consider:

  * How many people are in the workforce, and how many are projected to
    be working for the next 10 years. Build a model using:
    * Birth rate
    * Number of people unemployed
    * Death rate
  * How many people does the workforce have to sustain:
    * Number of Elderly - Calculate based on life span
    * Number of Young - Calculate based on when kids start working.
      This is a grey area, since kids work unofficially (by helping their
      parents, or even getting a job off-book).
    * Retirement age
  * Education:
    * Data at [](file://../owid-datasets/datasets/"Global Data Set on Education Quality (1965-2015) - Altinok, Angrist, and Patrinos (2018)"/"Global Data Set on Education Quality (1965-2015) - Altinok, Angrist, and Patrinos (2018).csv")
    * How many people can read and write.
      Reading/writing is just the first step. For a country to grow, people
      have to learn tools (like technology).
    * Higher education (college, and above). Some countries (ex. Germany)
      also have two tracks in college - one for the people who have better
      grades, and one for those that don't. Those who have better grades
      end up supervising those that don't.  Those that don't, become masters
      in a particular technology which they learn in college and are highly
      regarded as skilled workers.
    * Government support for education
  * Imports and exports
    * Imports into the country could be because there is more development
      (a plus) or because the population is growing (a minus).
      * Match import type to the development within the country. ex:
        * Iron ore is used for rebars for housing development. 
          Also check out other required resources (like cement for housing)
          to tell us what is going on in the country. Most countries
          use iron ore for housing - but some like US and Japan use wood.
        * Iron ore is also used to build pipes and machinery.
        * Educate ourselves on what inputs go into making:
          * Factories
          * Pipelines
          * Machinery
          * Clothes
          * Food (ex. fertilizer)
          * Water treatment plants
          * Roads, rail, ships for transport of goods
          
          If the items that are required are being imported, it will tell us 
          in which direction the country is growing.
    * Exports are great because they bring in outside money.
      Make sure natural resources are not being depleted due to exports.
      Exports also tell us what the country is capable of doing easily.
      All the resources needed to generate the product that is being exported,
      are easily available. Make sure those resources are enough for the
      forseeable future.
  * Energy expenditure
    * If more energy is being spent within the country, it's probably for
      factories - a plus if you're making your energy from sustainable
      resources. If not from sustainable resources, then you may continue
      for a while until you can do so. In the meantime, is the country moving
      to sustainable resources?
    * If energy is being brought in, is the country providing it able to
      maintain that export? Does the importing country need more each year,
      and does the exporting country able to support it.
  * Natural resources
    * Are natural resources plentiful, or are they being depleted?
    * Are they being used for the benefit of the population?
    * Create a list of natural resources that each country requires
  * Is the country able to feed it's own people?
  * Is drinking water available for the population?
  
## 1 Workforce

## 2 Education

```{r}
ed_quality <- read_csv("../owid-datasets/datasets/Global Data Set on Education Quality (1965-2015) - Altinok, Angrist, and Patrinos (2018)/Global Data Set on Education Quality (1965-2015) - Altinok, Angrist, and Patrinos (2018).csv")
ed_quality
```

We're going to ignore all lines for Entity names that have parentheses in them. They're for each area of the country. When we already have the country values, we don't need them. Also, they're sometimes only for some parts of the country (see US in Entity column).

We keep the current score and the slope of the scores. The current score tells us the latest education quality. It's the most important compared to the earlier scores. The slope of the scores tells us how much improvement there is in the country. Since the year column is 5 years apart, we can compare the slopes between countries.

```{r}
ed_quality <- ed_quality %>%
  select(Entity, Year, 
         "Average harmonised learning outcome score (Altinok, Angrist, and Patrinos (2018))") %>%
  rename(avg_score = "Average harmonised learning outcome score (Altinok, Angrist, and Patrinos (2018))",
         country = Entity) %>%
  filter(!str_detect(country, "\\(")) %>%
  group_by(country) %>%
  summarise(
    score_slope = lm(avg_score ~ Year)$coefficients[[2]],
    last_year = as.integer(max(Year)),
    avg_score = mean(avg_score)
  ) %>%
  select(country, last_year, avg_score, score_slope)
```

```{r}
ed_quality <- ed_quality %>%
  filter(!((country == "Belgium Flemish") |
           (country == "Belgium French")))
ed_quality
```

```{r}
ed_spending <- read_csv("../owid-datasets/datasets/Government Education Expenditure 1960-2010 - Szirmai (2015)/Government Education Expenditure 1960-2010 - Szirmai (2015).csv")
ed_spending
```

Rename the government spending column:
```{r}
ed_spending <- ed_spending %>%
  rename(country = Entity,
         govt_spend = "Government Education Expenditure 1960-2010 (Szirmai (2015))")
```

```{r}
ed_spending <- ed_spending %>%
  filter(!is.na(govt_spend)) %>%
  mutate(govt_spend = as.numeric(govt_spend)) %>%
  group_by(country) %>%
  summarise(
    last_year = as.integer(max(Year)),
    avg_govt_spend_perc_gdp = mean(govt_spend),
    slope_govt_spend_perc_gdp = lm(govt_spend ~ Year)$coefficients[[2]]
  )

ed_spending
```

Let's see what is the last year we have for countries' education-spending data:
```{r}
sapply(ed_spending, function(x) length(unique(x)))
```

So we have only `r length(ed_spending$entity)` countries. Let's see if they're right.
There are a few NA values in the slope. Let's remove those countries.
Also "High-income countries" is a row - let's remove it. 

```{r}
ed_spending <- ed_spending %>%
  filter(!is.na(avg_govt_spend_perc_gdp) &
           !is.na(slope_govt_spend_perc_gdp)) %>%
  filter(country != "High-income countries")

ed_spending
```
Now there are only `r length(ed_spending$entity)` countries. We may decide not to use this table, since it limits our options to only `r length(ed_spending$entity)` countries. Or, we could decide to use it if the countries we select fall into this group. Then we could compare them using this table as well. Let's keep the table for now.

## 3 Imports and Exports

Let's read in the dataset for exports and imports:
```{r}
flows <- 
  read_excel("../IRP Material Flows Database Bundle Post October 2020/UNEP IRP MFADB Vers 2018.xlsx",
             sheet = "13_Category")

flows
```
```{r}
unique(flows$Country)
```

Leave only countries in our dataframe - not regions.
```{r}
flows <- flows %>%
  filter(!(Country %in% c("Asia + Pacific", "EECCA",
                          "Latin America + Caribbean", "North America",
                          "West Asia", "World")))
flows
```

In this table:

  * DE  = Domestic Extraction. What the economy creates organically.
  * DMI = Direct Material Input. Measure of all material that is used in
          the economy (including extracted local and imported external material).
          Don't look at this in regions like the EU, since importing into an
          EU country will count double since the EU is consuming + importing
          the same amount.
          DMI = DE + IM
  * DMC = Domestic Material Consumption = Domestic extraction - exports + imports
          DMC = DE + IM - EX
  * PTB = Physical Trade Balance = Physical imports - physical exports
          PTB = IM - EX
  
So we're mainly interested in DMC. If this amount is growing compared to the
population, we can say the economy is consuming larger amounts of material,
and so it is growing. Notice that this is not related to monetary growth,
just the consumption habits of people.

We're also interested in PTB since it tells us if the economy is generating
more of the material it needs (negative PTB) or less (positive PTB).

```{r}
flows <- flows %>%
  filter((Flow == "DMC") | (Flow == "PTB")) %>%
  select(-FlowAndUnits) %>%
  rename(product = Product) %>%
  pivot_longer(`1970` : `2017`,
               names_to = "year",
               values_to = "amount")
flows <- flows %>%
  mutate(year = as.integer(year))

flows$product <- stringr::str_replace(flows$product, 
                     "Non-metallic minerals - construction dominant",
                     "construction minerals")
flows$product <- stringr::str_replace(flows$product,
    "Non-metallic minerals - industrial or agricultural dominant",
    "ind/ag minerals")

names(flows) <- stringr::str_to_lower(names(flows))

# Since there are only tonnes as units, drop the units column
unique(flows$units)
flows <- flows %>% select(-units)

flows
```
```{r}
flows %>%
  summarise(across(country:year, n_distinct))

unique(flows$product)
unique(flows$flow)
```

Let's look at DMC for Afghanistan:

```{r}
industrial_products <- 
  c("Coal", "Ferrous ores", "Natural Gas", "Non-ferrous ores", 
    "construction minerals",
    "ind/ag minerals",
    "Petroleum")

flows %>%
  filter(country == "Afghanistan") %>%
  filter(flow == "DMC") %>%
  filter(product %in% industrial_products) %>%
  ggplot(aes(year, amount)) +
  geom_line(aes(group = product, color = product), na.rm = TRUE) +
  scale_x_continuous(breaks = c(1970, 1990, 2010))
```

Let's break down the total amount of minerals and energy consumed by each
country per year

```{r}
energy <- c("Coal", "Natural Gas", "Petroleum")
minerals <- c("Ferrous ores", "Non-ferrous ores", "construction minerals",
              "ind/ag minerals")

flows <- flows %>%
  filter((product %in% energy) | (product %in% minerals)) %>%
  mutate(
    product = case_when(
      product %in% energy ~ "energy",
      product %in% minerals ~ "minerals"
    )
  ) %>%
  group_by(country, product, flow, year) %>%
  summarise(
    amount = sum(amount)
  )
flows
```

```{r}
all_na <- function(amount) {
  (n_distinct(amount) == 1) & is.na(amount[[1]])
}

DMC_PTB <- function(prod) {
  flows %>%
    filter(product == prod) %>%
    group_by(country, flow) %>%
    arrange(year) %>%
    summarise(
      last_year = max(year),
      new_amount = ifelse (all_na(amount), 
                           NA, 
                           mean(amount, na.rm = TRUE)),
      new_amount_slope = ifelse (all_na(amount), 
                                 NA, 
                                 lm(amount ~ year)$coefficients[[2]])
    ) %>%
    rename(
      amount = new_amount,
      amount_slope = new_amount_slope
    ) %>%
    mutate(product = prod) %>%
    select(country, last_year, product, flow, amount, amount_slope)
}

dmc_ptb_energy = DMC_PTB("energy")
dmc_ptb_energy
dmc_ptb_minerals = DMC_PTB("minerals")
dmc_ptb_minerals

dmc_ptb_flows <- dmc_ptb_energy %>% 
  full_join(dmc_ptb_minerals) %>%
  arrange(country, product, flow)
dmc_ptb_flows
```
Notice that these values are not compared to the population.
We cannot really compare these countries unless we compare consumption
per million people (for example). Otherwise, the denser countries will
always seem to be consuming more.

## Populations

```{r}
life_exp_and_pop <- read_csv("../owid-datasets/datasets/Life expectancy & population - Gapminder (2019), UN (2019), and Our World In Data (2019)/Life expectancy & population - Gapminder (2019), UN (2019), and Our World In Data (2019).csv")
life_exp_and_pop
```

Let's remove the rows with year = 1800 and year = 1950
```{r}
life_exp_and_pop <- life_exp_and_pop %>% filter((Year != 1800) & (Year != 1950))
names(life_exp_and_pop) <- stringr::str_to_lower(names(life_exp_and_pop))
life_exp_and_pop <- life_exp_and_pop %>% 
  rename(life_expectancy = `life expectancy`) %>%
  rename(country = entity)
life_exp_and_pop
```
Let's split the table into two - one with just life expectancy, and
one with just the population. This will allow us to use just the data we need.

```{r}
# Sanity check
sum(is.na(life_exp_and_pop))
```

```{r}
life_exp <- life_exp_and_pop %>%
  select(-population) %>%
  group_by(country) %>%
  summarise(
    last_year = max(year),
    life_exp = mean(life_expectancy, na.rm = TRUE),
    slope_life_exp = lm(life_expectancy ~ year)$coefficients[[2]]
  )

life_exp
```

To find population per age group, we can use another table from OWID:
```{r}
pop_per_age_group <- read.csv("../owid-datasets/datasets/Total population by broad age group, both sexes, 1950-2100 - UN Population Division (2015)/Total population by broad age group, both sexes, 1950-2100 - UN Population Division (2015).csv")
pop_per_age_group
```
Change column names to lowercase. Make names shorter.
```{r}
names(pop_per_age_group) <- stringr::str_to_lower(names(pop_per_age_group))
pop_per_age_group <- pop_per_age_group %>%
  rename(age_0_to_4 = `x0.to.4..un.population.division..2015..`) %>%
  rename(age_5_to_14 = `x5.to.14..un.population.division..2015..`) %>%
  rename(age_15_to_24 = `x15.to.24..un.population.division..2015..`) %>%
  rename(age_25_to_60 = `x25.to.60..un.population.division..2015..`) %>%
  rename(age_70_plus = `x70...un.population.division..2015..`) %>%
  rename(country = entity) %>%
  filter(year < 2015)  # Anything after 2014 is prediction, since data was
                       # collected in 2015. We don't want it.
pop_per_age_group
```
```{r}
pop_per_age_group <- pop_per_age_group %>%
  group_by(country, year) %>%
  summarise(
    year = year,
    age_0_to_4 = age_0_to_4,
    age_5_to_14 = age_5_to_14,
    age_15_to_24 = age_15_to_24,
    age_25_to_60 = age_25_to_60,
    age_70_plus = age_70_plus,
    total_pop = sum(age_0_to_4 : age_70_plus, na.rm = TRUE)
  ) %>%
  group_by(country) %>%
  summarise(
    last_year = last(year),
    age_0_to_4 = sum(age_0_to_4),
    age_5_to_14 = sum(age_5_to_14),
    age_15_to_24 = sum(age_15_to_24),
    age_25_to_60 = sum(age_25_to_60),
    age_70_plus = sum(age_70_plus),
    slope_total_pop = lm(total_pop ~ year)$coefficients[[2]],
    total_pop = sum(total_pop)
  ) %>%
  select(country, last_year, age_0_to_4, age_5_to_14, age_15_to_24,
         age_25_to_60, age_70_plus, total_pop, slope_total_pop)
```
```{r}
pop_per_age_group
```


## 4 Intrastructure development

## Load tables into Postgres

```{r}
db_connect <- function(db = NULL, host, port) {
  
  user <- keyring::key_list("LocalPostgresAccess")
  
  dbConnect(RPostgres::Postgres(),
            dbname = db,
            host = host,
            port = port,
            user = secrets$username,
            password = keyring::key_get("LocalPostgresAccess", 
                                        user$username))
}

db_disconnect <- function(con) {
  dbDisconnect(con)
}
```

```{r}
con <- db_connect("countries", "localhost", "5432")
```

```{r}
create_postgres_table <- function(con, table_name, table_df) {
  if (!dbExistsTable(con, table_name)) {
    dbCreateTable(con, table_name, table_df)
    dbAppendTable(con, table_name, table_df)
  }
}

remove_all_postgres_tables <- function(con) {
  dbRemoveTable(con, "ed_spending")
  dbRemoveTable(con, "ed_quality")
  dbRemoveTable(con, "life_exp")
  dbRemoveTable(con, "pop_per_age_group")
  dbRemoveTable(con, "dmc_ptb_flows")
}
remove_all_postgres_tables(con)

create_postgres_table(con, "ed_spending", ed_spending)
create_postgres_table(con, "ed_quality", ed_quality)
create_postgres_table(con, "life_exp", life_exp)
create_postgres_table(con, "pop_per_age_group", pop_per_age_group)
create_postgres_table(con, "dmc_ptb_flows", dmc_ptb_flows)
```
```{r}
dbListTables(con)
dbListFields(con, "ed_spending")

# Read small tables using dbReadTable
e_s <- dbReadTable(con, "ed_spending")
e_s

# Read small tables using dbGetQuery
e_s <- dbGetQuery(con, "select * from ed_spending")
e_s

# Read larger tables using dbSendQuery
# where query is for the entire table,
# and dbFetch with the batch size.
```

```{r}
# Runs a query "query" using the DBI library against the connection "con."
# Applies the "apply_func" function to each result-set before appending
# it to the table.
# Returns the table.
run_chunked_query <- function(con, query, n = 10, apply_func = identity) {
  
  result_set <- dbSendQuery(con, query)
  table_result <- NULL

  while(dbHasCompleted(result_set) == FALSE) {
    table_chunk <- apply_func(dbFetch(result_set, n = n))
    if (is.null(table_result)) {
      table_result <- table_chunk
    } else {
      table_result <- rbind(table_result, table_chunk)
    }
  }

  dbClearResult(result_set)
  table_result
}
```

```{r}
run_chunked_query(con, "select * from ed_spending", 10)
```

Useful DBI commands:

  * dbExistsTable(con, "trial_db")
  * dbWriteTable(con, "trial_db", trial_data): Creates table from df
  * dbWriteTable(con, "trial_db", trial2_data, overwrite = TRUE)
  * dbWriteTable(con, "trial_db", trial2_data, append = TRUE)
  * dbExecute(con, "INSERT into trial_db (x, y) VALUES (32, 'c'), (45,                                                         'k'), (61, 'h')")
    Inserts row into table. You can do the same with dbSendStatement
    and the same parameters as dbExecute
  * dbListTables(con)
  * dbRemoveTable(con, "trial_db)
  * dbQuoteIdentifier(con, "rating")
    Quote an identifier before pasting it into a SQL statement
  * dbQuoteLiteral(con, "G")
  
    Quote an literal before pasting it into a SQL statement
    query <- paste0("SELECT title, ", safe_id, 
                    " FROM film WHERE ", safe_id, " = ", safe_param )
    
  * Parametrizing queries:
    The "?" character in the query is replaced with the param value.
    
    q_params <- list('G', 90)
    query <- paste("SELECT title, rating, length FROM film WHERE 
                    rating = ? and length >= ?")
    res <- dbSendQuery(con, query, params = q_params)
    dbFetch(res, n = 3)
    
  * The same query can be used multiple times with different parameters:
  
    res <- dbSendQuery(con, "SELECT * FROM film WHERE rating = ?")
    dbBind(res, list("G"))
    dbFetch(res, n = 3)
    dbBind(res, list("PG"))
    dbFetch(res, n = 3)
    
  * The same query can be used at once with multiple values of the param:
  
    res <- dbSendQuery(con, "SELECT * FROM film WHERE rating = ?")
    dbBind(res, list(c("G", "PG")))
    dbFetch(res, n = 3)